---
title: "Group_7_Project"
author: "Riyan"
date: "2026-04-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Importing Libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stats)
library(car)
library(readr)
```

# Task 1

#### Task 1.1: Introduction to the Dataset:

***Diabetes problem*** : <br/> Diabetes is one of the most dangerous
diseases in the modern world. A study on diabetes in 2019 [1] estimates
that diabetes accounts for roughly 11% of adult deaths in globe. Not
only it is dangerous itself, type 2 diabetes, which accounts for 90% of
diabetes, could lead to the development of other dangerous diseases such
as Chronic kidney disease [2]. Smoking and drinking are considered two
risk factors of diabetes [3]. These factors usually come together in
adult life since an increase in one will lead to an increase in the
other [4]. While some sources confirm that smoking does have an impact
on diabetes [3][5], the impact of drinking and its joint impact with
smoking remains controversial[3].Thus, this study will examine the
impact of both smoking and drinking on diabetes. It asks a main research
question: “How do smoking and drinking impact the likelihood of
diabetes?”. <br/>

***Kaggle dataset*** : <br/> To answer the main research question, we
use the dataset "Diabetes Health Indicators Dataset"[6] from Kaggle. It
is collected by BRFSS and published on Kaggle by Alex Teboul. It is the
collection of more than 400,000 survey for American people in 2015. The
dataset [6] has *CC0: Public Domain* license which mean it can be
analyzed without permission [7]. It contains information about
individuals’ health conditions, lifestyle behaviors, and demographic
characteristics. It includes variables such as BMI, smoking status,
alcohol consumption, physical activity, and diabetes status. The dataset
is structured in a tabular format where each row represents an
individual and each column represents a specific health-related
variable.<br/>

***Reason for the choice of the dataset*** :<br/>This data set was
chosen because it is directly related to the research topic and includes
important variables such as smoking and drinking for our research.
Furthermore, [6] is one of reliable not synthetic dataset in Kaggle. The
large size of the dataset also helps in producing reliable and
meaningful results.

Loading Dataset

```{r}

data <- read_csv("./archive/diabetes_binary_health_indicators_BRFSS2015.csv")

head(data)
str(data)
```

#### Task 1.2: Data Summarization:

```{r}
#Summary Statistics
#summary
#We are not finding Mode as data is continuous

numeric_data <- data %>% select(where(is.numeric))

descriptive_stats <- data.frame(
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) median(x, na.rm = TRUE)),
  Q1 = sapply(numeric_data, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Q3 = sapply(numeric_data, function(x) quantile(x, 0.75, na.rm = TRUE)),
  IQR = sapply(numeric_data, function(x) IQR(x, na.rm = TRUE))
)

descriptive_stats

```

```{r}
# Relative Frequency Tables
print("HighChol Distribution (%)")
prop.table(table(data$HighChol)) * 100

print("CholCheck Distribution (%)")
prop.table(table(data$CholCheck)) * 100

print("Stroke Distribution (%)")
prop.table(table(data$Stroke)) * 100

print("Heart Disease or Attack Distribution (%)")
prop.table(table(data$HeartDiseaseorAttack)) * 100

print("Physical Activity Distribution (%)")
prop.table(table(data$PhysActivity)) * 100

print("Fruits Consumption Distribution (%)")
prop.table(table(data$Fruits)) * 100

print("Vegetables Consumption Distribution (%)")
prop.table(table(data$Veggies)) * 100

print("Heavy Alcohol Consumption Distribution (%)")
prop.table(table(data$HvyAlcoholConsump)) * 100

print("Any Healthcare Coverage Distribution (%)")
prop.table(table(data$AnyHealthcare)) * 100

print("Diabetes Distribution (%)")
prop.table(table(data$Diabetes_binary)) * 100

print("Smoking Distribution (%)")
prop.table(table(data$Smoker)) * 100

print("Drinking Distribution (%)")
prop.table(table(data$HvyAlcoholConsump)) * 100
```

```{r}
#Data Types Conversion
data$Diabetes_binary <- as.factor(data$Diabetes_binary)
data$Sex <- as.factor(data$Sex)
data$HighBP <- as.factor(data$HighBP)
data$Smoker <- as.factor(data$Smoker)
data$HvyAlcoholConsump <- as.factor(data$HvyAlcoholConsump)

```

#### Task 1.3: Data Visualization:

```{r}
#Histogram
ggplot(data, aes(x = BMI)) +
  geom_histogram(bins = 30) +
  ggtitle("Distribution of BMI")
```

```{r}
#Boxplot
ggplot(data, aes(x = Diabetes_binary, y = BMI)) +
  geom_boxplot() +
  ggtitle("BMI vs Diabetes Status")
```

```{r}
#Bar Chart
ggplot(data, aes(x = Diabetes_binary)) +
  geom_bar() +
  ggtitle("Diabetes Distribution")
```

```{r}
ggplot(data, aes(x = Smoker)) +
  geom_bar() +
  ggtitle("Smoking Distribution")
```

```{r}
ggplot(data, aes(x = HvyAlcoholConsump)) +
  geom_bar() +
  ggtitle("Drinking Distribution")
```

#### Task 1.4 : Data Quality Assessment, Data Wrangling and Cleaning:

Let's first check for any missing values:

```{r}
#Checking for missing values
colSums(is.na(data))
```

As there are no missing values, we can move ahead to remove any outliers
in our dataset, so from histogram we created, we found out anything less
than 10 and greater than 60 can be considered as an outlier

```{r}
#Removing Outliers
data <- data %>% filter(BMI > 10 & BMI < 60)
show(head(data))
```

# Task 2

#### Task 2.1: Research Question Development:

As stated in Task 1, the main research question for this study is: *“How
do smoking and heavy alcohol consumption impact the likelihood of
diabetes?”*. In order to answer this, two sub-questions are asked
<br/><br/> (a). Does Smoking or Drinking individually impact the
likelihood of diabetes?<br/><br/> (b). How do Smoking and Drinking
jointly impact the likelihood of diabetes?<br/> <br/> In Task 2, we
focus on answer the question (a) by using chi-square test. (a) can be
further divided into two sub-questions (a.1) Does Smoking impact the
likelihood of diabetes? (a.2) Does Drinking impact the likelihood of
diabetes?

The variables used are :

-   Smoker (0 = No, 1 = Yes):Have you smoked at least 100 cigarettes in
    your entire life?<br/><br/>

-   HvyAlcoholConsump (0 = No, 1 = Yes) : Heavy drinkers where adult men
    having more than 14 drinks per week or adult women having more than
    7 drinks per week<br/><br/>

-   Diabetes_binary (0 = No diabetes, 1 = PreDiabetes or Diabetes):
    <br/><br/>

All these variables are discrete between 0 and 1, so chi-square test is
suitable.

#### Task 2.2: Data Preparation:

For this part, the dataset was prepared by selecting the required
variables. These variables were already cleaned and converted earlier.
<br/>

Since all variables are categorical, contingency tables were created to
observe how diabetes levels are distributed across smoking and alcohol
groups.

```{r}
print("Smoking vs Diabetes")
table(data$Smoker, data$Diabetes_binary)

print("Alcohol vs Diabetes")
table(data$HvyAlcoholConsump, data$Diabetes_binary)

```

***Checking assumptions for chi-square test*** :<br/>

-   ***1] All variables used in the test should be categorical:*** In
    this case, Smoking, Alcohol consumption, and Diabetes are
    categorical variables, so this condition is satisfied. <br/>

-   ***2] The observations should be independent:*** Since each row in
    the dataset represents a different individual, this assumption is
    also satisfied. <br/>

-   ***3] The expected frequency in each cell should be at least 5***:
    That is given below:

```{r}
print("Expected frequency for Smoking")
chisq.test(table(data$Smoker, data$Diabetes_binary))$expected
print("Expected frequency for Drinking")
chisq.test(table(data$HvyAlcoholConsump, data$Diabetes_binary))$expected
```

Those assumptions are satisfied so we can conduct our test.

#### Task 2.3: Hypothesis construction and testing:

***Question (a.1): Does Smoking impact the likelihood of
diabetes?***<br/>

-   ***Null H0***: Smoking and diabetes are not related<br/>

-   ***Alternative H1***: Smoking and diabetes are related

```{r}
chisq.test(table(data$Smoker, data$Diabetes_binary))
```

***Question (a.2): Does Drinking impact the likelihood of
diabetes?***<br/>

-   ***Null H0***: Alcohol consumption and diabetes are not related
    <br/>

-   ***Alternative H1***: Alcohol consumption and diabetes are related

```{r}
chisq.test(table(data$HvyAlcoholConsump, data$Diabetes_binary))
```

Result: The p-values for both tests are less than 0.05, so the null
hypothesis is rejected. This means that both smoking and alcohol
consumption have a statistically significant association with diabetes
status, including no diabetes and diabetes.

#### Task 2.4: Rationale for test Choice:

The chi-square test is used because all variables are categorical. It
helps in checking whether there is a relationship between two
categorical variables. One limitation is that it only shows association
and not how strong the relationship is. Also, it does not include other
factors.

# Task 3:

#### Task 3.1: Research Question:

In Task 2, the impact of smoking and alcohol consumption on diabetes was
analyzed separately using chi-square test. That helped in understanding
their individual effects.<br/>

In this task, the focus is on understanding their combined impact. We
answer the question (b). How do Smoking and Drinking jointly impact the
likelihood of diabetes?<br/>

This is important because in real life, people are often exposed to
multiple lifestyle factors together. Furthermore, as stated in [3], the
past literature usually look at individual impacts such as what we did
in task 2, but their joint impact is not well-studied. So instead of
just looking individual impacts like task 2, Task 3 use advance
analytical approach by using Multivarte Logistic Regression to study
their join impact on the likelihood of having diabetes.<br/>

#### Task 3.2: Model Selection:

[8] taught us that when dealing with categorical variables, we should
use Multivariate Logistic Regression. It can be used in R by using
Generalized Linear Model ***glm***

The independent variables are:

Smoker HvyAlcoholConsump

Before applying the model, I converted all variables into factors (even
thugh we already did).

```{r}
# Convert variables to factors
data$Diabetes_binary <- as.factor(data$Diabetes_binary)
data$Smoker <- as.factor(data$Smoker)
data$HvyAlcoholConsump <- as.factor(data$HvyAlcoholConsump)
```
***Checking assumptions for Multivariate Logistic Regression*** :<br/>

-   ***1] The dependent variable should be categorical***: In this case,
    Diabetes is categorical, so this condition is satisfied. <br/>

-   ***2] The observations should be independent*** Each row represents
    a different individual, so this assumption is satisfied. <br/>

-   ***3] There should not be a strong relationship between independent
    variables***: Since only Smoking and Alcohol are used, this issue is
    minimal.

Those assumptions are satisfied so the regression model will be created
to study the joint effect of smoking and alcohol on diabetes.
#### Task 3.3: Model Fitting and Evaluation:

```{r}
model <- glm(Diabetes_binary ~ Smoker + HvyAlcoholConsump, data = data, family = binomial)

summary(model)
```

**Result:** Smoking has a positive coefficient, which means smoking
increases the likelihood of diabetes. Alcohol has a negative
coefficient(-0.986419) in this dataset, which shows an inverse
relationship. Both has pretty small p value which mean that they are
signficiant and doesn't happen by chance. However, this is show no
different than task 2. [8] show that we can dive deep into the logistic
model by using the odd ratios:

```{r}
exp(coef(model))
```

To Intepret this: - Smoker: for every unit increase in Smoker1, the
likelihood of having diabetes increase by 47,47% <br/> -
HvyAlcoholConsump1: for every unit increase in HvyAlcoholConsump1, the
likelihood of having diabetes decrease by (1- 0.3387765) \* 100% \~
66,12% <br/>

#### Task 3.4: Insight Extraction:

The result of 3.3 is quite suprising result since the impact of Alcohol
is larger than the impact of Smoking in reverse direction. Which might
indicate that Drinking cancel the consequence of Smoking in term of
diabetes.<br/>

To better understand the combined effect of smoking and alcohol on
diabetes, we can use simple summaries and visualization.

```{r}
# Create groups
data$group <- interaction(data$Smoker, data$HvyAlcoholConsump)

# Rename groups
levels(data$group) <- c(
  "Non-Smoker & Non-Drinker",
  "Smoker & Non-Drinker",
  "Non-Smoker & Drinker",
  "Smoker & Drinker"
)

# Average diabetes level
tapply(as.numeric(as.character(data$Diabetes_binary)), data$group, mean)
```

Visualization:

```{r}
library(ggplot2)

data$Diabetes_binary <- factor(data$Diabetes_binary,
                            levels = c(0,1),
                            labels = c("No Diabetes","Diabetes"))

# Plot
ggplot(data, aes(x = group, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes Distribution by Smoking and Alcohol",
       x = "Group",
       y = "Proportion") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

As seen above, without Drinking, the Diabetes propotion in both
Non-Smoking and Smoking is higher compared to with Drinking.

# Task 4

#### Task 4.1 Result Discussion:

In Task 2, we have checked the individual impacts of drinking using chi
square test for contigency table. The result of both cases show a small
p value \< 0.5 which indicate that both have strong respective impact on
the likelihood of having diabetes. However, whether they together
increase the likelihood or what is unclear. Thus, in task 3 we examine
the joint impact both of them by introducing and using Multivate
Logistic Regression. Logistic not Linear because our respond variable is
categorical. The result show that both variable Drinking and Smoking
have opposite impacts. Smoking increases the chance of having diabetes
whereas Drinking decreases it. What make our study standout is because
we found that the impact of Drinking 66,12% is stronger than that of
Smoking 47,47%. Even though it has good impact on diabetes, the study
doesn't endorse the heavy use of Alcohol to cancel impact of Smoking
because Alcohol itself could lead to many other risks. As [3] suggests,
we could take advantage of this impact by consuming Alcohol with
adequate quantity.

#### Task 4.2 Limitations of the Analysis:

-   Small number of used variable: This dataset consists of many other
    factors such as BMI or Cholestorol,... up to 21 variables.However
    our study only involve two factors which are drinking and
    smoking.<br/>

-   Pre-diabetes: Our study group diabetes and pre-diabetes as a single
    group to study about the general diabetes.While this helps in term
    of simplicity, more complex pattern and insights might require the
    seperation. <br/>

-   Survey: Even though this is a collection from reliable organization.
    It is still survey-based which mean that they could make mistake
    while filling the survey

#### Task 4.3 Improvenments:

-   Actual Record: Instead of having, it might be better to ask the
    health centre for the actual record. This is better than the
    patients mannually fill the survey due to the strict procedure in
    Health Centre. However, it might raise some questions about the
    permission and ethics<br/>
-   More variables: More variables used in the study could help us in
    understanding the diabetes more<br/>
-   Separation for pre-diabetes: By adding the pre-diabetes, we could
    ask more questions. For example, we could further examine whether
    Smoking and Drinking leads to pre-diabetes more or directly to
    diabetes.

# Reference

[1] P. Saeedi et al., “Mortality attributable to diabetes in 20–79 years
old adults, 2019 estimates: Results from the International Diabetes
Federation Diabetes Atlas, 9th edition,” Diabetes Research and Clinical
Practice, vol. 162, Art. no. 108086, 2020, doi:
10.1016/j.diabres.2020.108086.

[2] D. Xie et al., “Global burden and influencing factors of chronic
kidney disease due to type 2 diabetes in adults aged 20–59 years,
1990–2019,” Scientific Reports, vol. 13, no. 1, Art. no. 20234, 2023,
doi: 10.1038/s41598-023-47091-y.

[3] I. Paz-Graniel et al., “Alcohol, smoking, and their synergy as risk
factors for incident type 2 diabetes,” American Journal of Preventive
Medicine, vol. 69, no. 5, Art. no. 108011, 2025, doi:
10.1016/j.amepre.2025.108011.

[4] Y. R. Yanjun, C. B. Castro, and J.-P. L., “Drink and smoke; drink or
smoke? The interdependence between alcohol and cigarette consumption for
men in China,” Empirical Economics, vol. 58, no. 3, pp. 921–955, 2020,
doi: 10.1007/s00181-018-1557-x.

[5] Health Canada, “Tobacco and type 2 diabetes,” Jul. 26, 2023.
[Online]. Available:
<https://www.canada.ca/en/health-canada/services/health-concerns/tobacco/legislation/tobacco-product-labelling/tobacco-type-2-diabetes.html>

[6] A. Teboul, “Diabetes Health Indicators Dataset,” Kaggle, n.d.
[Online]. Available:
<https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data>
<br/>

[7] Creative Commons, “Creative Commons — CC0 1.0 Universal,” 2019.
[Online]. Available:
<https://creativecommons.org/publicdomain/zero/1.0/> <br/>

[8] U. Riswanto, “A Beginner’s Guide to Multivariate Logistic Regression
in R,” Medium, Apr. 20, 2025. [Online]. Available:
<https://ujangriswanto08.medium.com/a-beginners-guide-to-multivariate-logistic-regression-in-r-9aa3b763f564>
<br/>
